Brother, Can You Spare a Manufacturing Job? How Voters React to Deindustrialization

### 2021-12-12 Final Draft/Final Draft.Rmd
updates

---
title: "Brother, Can You Spare a Manufacturing Job?  How Voters React to Deindustrialization"
short: "Brother, Can You Spare a Manufacturing Job? "
month: "`r format(Sys.Date(), '%m')`"
year: "`r format(Sys.Date(), '%Y')`"
author:
  - name: Catherine Darin
    firstname: Catherine
    surname: Darin
    email: cdarin@hks.harvard.edu
    affiliation: Harvard University
  - name: Zagreb Mukerjee
    firstname: Zagreb
    surname: Mukerjee
    email: zagrebmukerjee@fas.harvard.edu
    affiliation: Harvard University
acknowledgements: | 
  Code available at https://github.com/zagrebmukerjee/ReplicationPaper
abstract: |
  What led to Donald Trump's surprising 2016 election victory? This paper examines the potential contribution of deindustrialization. We exploit county-level differences in manufacturing density as a source of exogenous variation, using both instrumental-variable regression and covariate balancing methods. Counter-intuitively, we find that counties that gained manufacturing jobs from 2012-2015 are more likely to have swung towards Trump. We believe this is due to an omitted variable - the job loss experienced from 2004 to 2012. Counties that experienced prior deindustrialization were more likely to exhibit increased support for Trump, as well as a slight manufacturing rebound in 2012-2015. Dis-aggregating this result by race shows that the Democratic vote share fell where layoffs affected white populations, and rose where layoffs affected nonwhites. This suggests a racial component to how voters process economic hardship.
output: rticles::aea_article
header-includes: 
 - \usepackage{graphicx}
 - \usepackage{amsmath}
 - \usepackage{amssymb}
 - \usepackage{placeins}
---

@@

Like \cite{Baccini21}, our results suggest that the experience of deindustralization was associated with support for Donald Trump, particularly among whites. However, our results differ from \cite{Baccini21} in regards to the *timing* of the impact of manufacturing job losses on employment. In fact, using our preferred measure of deindustrialization - *net* change in manufacturing employment - we find that counties with manufacturing job *gains* from 2012-2015 tended to swing towards Donald Trump in 2016. Given this counter-intuitive result, we extend the analysis to the 2004-2015 period in order to capture the long-term effects of deindustrialization on a region. Doing so, we find that regions experiencing more manufacturing job losses from 2004-2015 swung towards Donald Trump in 2016, and that this occurred more when those losing jobs were white. We get similar result using our covariate-matching approach. Relative to \cite{Baccini21}, our results point to the substantive importance of the timing of manufacturing layoffs on political outcomes.

@@

Finally, it is important to note that, in the counterfactual and simulation exercises shown in Figures \ref{counterfactual1}, \ref{counterfactual2}, and \ref{counterfactual3}, we estimate the change in Democratic vote share assuming various levels of white manufacturing job losses *holding all else equal, including non-white manufacturing job losses.* Given that white and non-white manufacturing job losses are weakly-to-moderately correlated across counties (r = .33), this is likely a poor assumption, because, in reality, a decline in white manufacturing job losses would likely correspond to a decline in non-white manufacturing job losses. Given that non-white manufacturing job losses were expected to *increase* Democratic vote share in 2016, this means that our simulation results may be overestimating effect sizes because it does not take into account this second-order impact.

@@

```
\end{figure}
\FloatBarrier
\subsubsection{Timing}

@@

ntile_summary <- ntile_summary %>% mutate(n_tile_04_11 = as.numeric(n_tile_04_11))
# plot 2004-2011 manu layoffs by deciles
ggplot(data = ntile_summary,
       aes(x = n_tile_04_11, y = mean_04_11)) +
  theme_minimal() +
  ylab("Manufacturing Losses (04-11)") +
  xlab("Deciles")+
  ggtitle("Manufacturing Job Losses from 2004-2011 by Deciles") +
  geom_col(fill = "#4287f5") +
  scale_x_continuous(breaks=seq(1,10,1))

```
\begin{figurenotes}
Deciles are determined based on our preferred measure of deindustrialization: total net manufacturing job losses from 2004-2015 divided by total 2004 employment. Counties in decile 1 had the lowest manufacturing job losses (i.e., an increase in manufacturing employment) while counties in decile 10 had the highest manufacturing job losses. The y-axis shows the mean measure of manufacturing job lossess (2004-2015) per 2004 worker across counties in each decile group.
\end{figurenotes}
\end{figure}

@@

# plot 2012-2015 manu layoffs by deciles
ggplot(data = ntile_summary,
       aes(x = n_tile_04_11, y = mean_12_15)) +
  theme_minimal() +
  ylab("Manufacturing Layoffs (12-15)") +
  xlab("Decile of Manufacturing Layoffs (04-11)")+
  ggtitle("Manufacturing Layoffs from 2012-2015 by 2004-2011 Deciles") +
  geom_col(fill = "#4287f5") +
  scale_x_continuous(breaks=seq(1,10,1))

>>>>

---
title: "Brother, Can You Spare a Manufacturing Job?  How Voters React to Deindustrialization"
short: "Brother, Can You Spare a Manufacturing Job? "
month: "`r format(Sys.Date(), '%m')`"
year: "`r format(Sys.Date(), '%Y')`"
author:
  - name: Catherine Darin
    firstname: Catherine
    surname: Darin
    email: cdarin@hks.harvard.edu
    affiliation: Harvard University
  - name: Zagreb Mukerjee
    firstname: Zagreb
    surname: Mukerjee
    email: zagrebmukerjee@fas.harvard.edu
    affiliation: Harvard University
acknowledgements: | 
  Code available at https://github.com/zagrebmukerjee/ReplicationPaper
abstract: |
  What led to Donald Trump's surprising 2016 election victory? This paper examines the potential contribution of deindustrialization. We exploit county-level differences in manufacturing density as a source of exogenous variation, using both instrumental-variable regression and covariate balancing methods. Counter-intuitively, we find that counties that gained manufacturing jobs from 2012-2015 are more likely to have swung towards Trump. We believe this is due to an omitted variable - the job loss experienced from 2004 to 2012. Counties that experienced prior deindustrialization were more likely to exhibit increased support for Trump, as well as a slight manufacturing rebound in 2012-2015. Dis-aggregating this result by race shows that the Democratic vote share fell where job losses affected white populations, and rose where job losses affected nonwhites. This suggests a racial component to how voters process economic hardship.
output: rticles::aea_article
header-includes: 
 - \usepackage{graphicx}
 - \usepackage{amsmath}
 - \usepackage{amssymb}
 - \usepackage{placeins}
---

@@

Like \cite{Baccini21}, our results suggest that the experience of deindustralization was associated with support for Donald Trump, particularly among whites. However, our results differ from \cite{Baccini21} in regards to the *timing* of the impact of manufacturing job losses on employment. In fact, using our preferred measure of deindustrialization - *net* change in manufacturing employment - we find that counties with manufacturing job *gains* from 2012-2015 tended to swing towards Donald Trump in 2016. Given this counter-intuitive result, we extend the analysis to the 2004-2015 period in order to capture the long-term effects of deindustrialization on a region. Doing so, we find that regions experiencing more manufacturing job losses from 2004-2015 swung towards Donald Trump in 2016, and that this occurred more when those losing jobs were white. We get similar result using our covariate-matching approach. Relative to \cite{Baccini21}, our results point to the substantive importance of the timing of manufacturing job losses on political outcomes.

@@

It is important to note that, in the counterfactual and simulation exercises shown in Figures \ref{counterfactual1}, \ref{counterfactual2}, and \ref{counterfactual3}, we estimate the change in Democratic vote share assuming various levels of white manufacturing job losses *holding all else equal, including non-white manufacturing job losses.* Given that white and non-white manufacturing job losses are weakly-to-moderately correlated across counties (r = .33), this is likely a poor assumption, because, in reality, a decline in white manufacturing job losses would likely correspond to a decline in non-white manufacturing job losses. Given that non-white manufacturing job losses were expected to *increase* Democratic vote share in 2016, this means that our simulation results may be overestimating effect sizes because it does not take into account this second-order impact.

@@

More broadly, we would like to acknowledge our focus on white manufacturing job losses over non-white manufacturing job losses. We focus on white manufacturing job losses for two reasons. First, in the United States, manufacturing employment and job losses have historically been predominately white (See Table \ref{tab:descTable}). Second, our regression results (See Table \ref{regResult04}) suggest that white manufacturing job losses have a larger negative effect on Democratic vote share relative to the magnitude of the positive effect from non-white manufacturing job losses. For these reasons, our results suggest that, in aggregate, white manufacturing job losses have had larger effect on political outcomes than non-white manufacturing job losses. However, we would like to highlight that the political consequences from non-white manufacturing job losses is a topic that is under-explored in this paper, and is an important area for future work.

\subsubsection{Timing}

@@

ntile_summary <- ntile_summary %>% mutate(n_tile_04_11 = as.numeric(n_tile_04_11))
# plot 2004-2011 manu layoffs by deciles
ggplot(data = ntile_summary,
       aes(x = n_tile_04_11, y = mean_04_11)) +
  theme_minimal() +
  ylab("Mean Manufacturing Job Losses (04-11)") +
  xlab("Decile of Manufacturing Job Losses (04-11)")+
  ggtitle("") +
  geom_col(fill = "#4287f5") +
  scale_x_continuous(breaks=seq(1,10,1))  + theme(text = element_text(size = 8))

```
\begin{figurenotes}
Deciles are determined based on our preferred measure of deindustrialization: total net manufacturing job losses from 2004-2015 divided by total 2004 employment. Counties in decile 1 had the lowest manufacturing job losses (i.e., an increase in manufacturing employment) while counties in decile 10 had the highest manufacturing job losses. The y-axis shows the mean measure of manufacturing job losses (2004-2015) per 2004 worker across counties in each decile group.
\end{figurenotes}
\end{figure}

@@

# plot 2012-2015 manu layoffs by deciles
ggplot(data = ntile_summary,
       aes(x = n_tile_04_11, y = mean_12_15)) +
  theme_minimal() +
  ylab("Mean Manufacturing Job Losses (12-15)") +
  xlab("Decile of Manufacturing Job Losses (04-11)")+
  ggtitle("") +
  geom_col(fill = "#4287f5") +
  scale_x_continuous(breaks=seq(1,10,1)) + theme(text = element_text(size = 8))

### 2021-12-12 Final Draft/Final-Draft.tex
updates

\begin{abstract}
What led to Donald Trump's surprising 2016 election victory? This paper
examines the potential contribution of deindustrialization. We exploit
county-level differences in manufacturing density as a source of
exogenous variation, using both instrumental-variable regression and
covariate balancing methods. Counterintuitively, we find that counties
that gained manufacturing jobs from 2012-2015 are more likely to have
swung towards Trump. We believe this is due to an omitted variable - the
job loss experienced from 2004 to 2012. Counties that experienced prior
deindustrialization were more likely to exhibit increased support for
Trump, as well as a slight manufacturing rebound in 2012-2015.
Disaggregating this result by race shows that the Democratic vote share
fell where layoffs affected white populations, and rose where layoffs
affected nonwhites. This suggests a racial component to how voters
process economic hardship.
\end{abstract}

@@

We also demonstrate uncertainity around our coefficient estimates using
a simulation appraoch. Specifically, for each percentile of white
manufacturing job losses, we simulate 1,000 election outcomes by drawing
from the distribution around our coefficent point estimate 1,000 times
(again, using the results from Model 3 in Table \ref{regResult04}). In
Figure \ref{counterfactual2}, we find that, assuming the 24th percentile
or lower of white manufacturing job losses, we find that the median
estimate in our simulation shows Democratic winning at least 270
electoral college votes. Using the 95\% confidence intervals produced
from the simulation exercise, we find that we would be 95\% confident
that Democrats would have won the national election had white
manufacturing job losses been at the 5th percentile or lower. This
result suggest that manufacturing job losses would have had to have been
substantially lower and \emph{negative} (e.g., job gains) in order for
the national election outcome to have been flipped with 95\% confidence.

@@

Using the same simulation approach described above, in Figure
\ref{counterfactual3} we illustrate the freuqency with which Demcorats
would have won the presidential election (270 votes) given each
percentile of white manufacturing job losses. Our simulation results
show that, when we assumed 24th percentile or lower of white
manufacturing job losses, Democrats won the presidential election in at
least 50 percent of simulations. When we assumed the 17th percentile or
lower, Democrats won the presidential election in at least 75 percent of
simulations. Taken together, these results suggest that Democrats would
have had a significantly higher chance of winning the presidential
election had white manufacturing job layoffs been substantially lower.

Finally, it is important to note that, in the counterfactual and
simulation exercises shown in Figures \ref{counterfactual1},
\ref{counterfactual2}, and \ref{counterfactual3}, we estimate the change
in Democratic vote share assuming various levels of white manufacturing
job losses \emph{holding all else equal, including non-white
manufacturing job losses.} Given that white and non-white manufacturing
job losses are weakly-to-moderately correlated across counties (r =
.33), this is likely a poor assumption, because, in reality, a decline
in white manufacturing job losses would likely correspond to a decline
in non-white manufacturing job losses. Given that non-white
manufacturing job losses were expected to \emph{increase} Democratic
vote share in 2016, this means that our simulation results may be
overestimating effect sizes because it does not take into account this
second-order impact.

@@

\FloatBarrier
\end{figure}
\FloatBarrier
\subsubsection{Timing}

@@

\begin{figure} \label{vShapePlot}
\caption{V-Shaped Recovery in Manufacturing Job Losses}

@@

\FloatBarrier
\begin{figurenotes}
"Stuff here"
\end{figurenotes}
\end{figure}

\begin{figure} \label{vShapePlot}
\caption{V-Shaped Recovery in Manufacturing Job Losses}

>>>>

\begin{abstract}
What led to Donald Trump's surprising 2016 election victory? This paper
examines the potential contribution of deindustrialization. We exploit
county-level differences in manufacturing density as a source of
exogenous variation, using both instrumental-variable regression and
covariate balancing methods. Counter-intuitively, we find that counties
that gained manufacturing jobs from 2012-2015 are more likely to have
swung towards Trump. We believe this is due to an omitted variable - the
job loss experienced from 2004 to 2012. Counties that experienced prior
deindustrialization were more likely to exhibit increased support for
Trump, as well as a slight manufacturing rebound in 2012-2015.
Dis-aggregating this result by race shows that the Democratic vote share
fell where layoffs affected white populations, and rose where layoffs
affected nonwhites. This suggests a racial component to how voters
process economic hardship.
\end{abstract}

@@

We also demonstrate uncertainty around our coefficient estimates using a
simulation approach. Specifically, for each percentile of white
manufacturing job losses, we simulate 1,000 election outcomes by drawing
from the distribution around our coefficient point estimate 1,000 times
(again, using the results from Model 3 in Table \ref{regResult04}). In
Figure \ref{counterfactual2}, we find that, assuming the 24th percentile
or lower of white manufacturing job losses, we find that the median
estimate in our simulation shows Democratic winning at least 270
electoral college votes. Using the 95\% confidence intervals produced
from the simulation exercise, we find that we would be 95\% confident
that Democrats would have won the national election had white
manufacturing job losses been at the 5th percentile or lower. This
result suggest that manufacturing job losses would have had to have been
substantially lower and \emph{negative} (e.g., job gains) in order for
the national election outcome to have been flipped with 95\% confidence.

@@

Using the same simulation approach described above, in Figure
\ref{counterfactual3} we illustrate the frequency with which Democrats
would have won the presidential election (270 votes) given each
percentile of white manufacturing job losses. Our simulation results
show that, when we assumed 24th percentile or lower of white
manufacturing job losses, Democrats won the presidential election in at
least 50 percent of simulations. When we assumed the 17th percentile or
lower, Democrats won the presidential election in at least 75 percent of
simulations. Taken together, these results suggest that Democrats would
have had a significantly higher chance of winning the presidential
election had white manufacturing job layoffs been substantially lower.

It is important to note that, in the counterfactual and simulation
exercises shown in Figures \ref{counterfactual1}, \ref{counterfactual2},
and \ref{counterfactual3}, we estimate the change in Democratic vote
share assuming various levels of white manufacturing job losses
\emph{holding all else equal, including non-white manufacturing job
losses.} Given that white and non-white manufacturing job losses are
weakly-to-moderately correlated across counties (r = .33), this is
likely a poor assumption, because, in reality, a decline in white
manufacturing job losses would likely correspond to a decline in
non-white manufacturing job losses. Given that non-white manufacturing
job losses were expected to \emph{increase} Democratic vote share in
2016, this means that our simulation results may be overestimating
effect sizes because it does not take into account this second-order
impact.

@@

Finally, we would like to acknowledge our focus on white manufacturing
job losses over non-white manufacturing job losses. In this analysis, we
focus on white manufacturing job losses for two reasons. First, in the
United States, manufacturing employment and job losses have historically
been predominately white (See Table \ref{tab:descTable}). Second, our
results suggest that the white manufacturing job losses have a larger
negative effect on Democratic vote share relative to the magnitude of
the positive effect from non-white manufacturing losses. For these
reasons, our results suggest that, in aggregate, white-manufacturing
losses have had a more significant effect on political outcomes than
non-white manufacturing job losses. However, we would like to highlight
that the political consequences from non-white deindustrialization is an
important area for future work.

\subsubsection{Timing}

@@

\begin{figure} \label{vShapePlot1}
\caption{V-Shaped Recovery in Manufacturing Job Losses: Manufacturing Job Losses (2004-2011) by (2004-2011) Decile Groupings}

@@

\FloatBarrier
\begin{figurenotes}
Deciles are determined based on our preferred measure of deindustrialization: total net manufacturing job losses from 2004-2015 divided by total 2004 employment. Counties in decile 1 had the lowest manufacturing job losses (i.e., an increase in manufacturing employment) while counties in decile 10 had the highest manufacturing job losses. The y-axis shows the mean measure of manufacturing job losses (2004-2015) per 2004 worker across counties in each decile group.
\end{figurenotes}
\end{figure}

\begin{figure} \label{vShapePlot2}
\caption{V-Shaped Recovery in Manufacturing Job Losses: Manufacturing Job Losses (2012-2015) by 2004-2011 Decile Groupings}
